Linear stability of magnetized massive protoplanetary disks

### 2014-03-31 paper/results.tex
tex updates

\section{MRI in self-gravitating disks}
In this section we focus on the MRI. We use the vertical field setup
of case 1. We first consider MRI modes with negligble densit/potential
perturbations to see the effect of self-gravity on the MRI through the
background stratification, then go on to consider MRI modes with
significant density/potential perturbations in massive disks.

\subsection{Influence of self-gravity on the MRI through the
  background equilibrium}

Here we use polytropic disks, which have a well-defined
thickness. The upper disk boundary is set to 
$Z_s=0.99H$. We fix $\beta=100$ and $k_xH=0.1$ unless otherwise
stated.

@@

\subsubsection{Uniform
  resistivity}  
We first calculate the MRI as a function of $Q$ with uniform
resistivity ($A=1$) for mid-plane Elsasser numbers $\Lambda_0\in[0.3,10]$.

Fig. \ref{compare_growth_poly_uniresis} plots the MRI growth rates as
a function of $Q$ and $\Lambda_0$. As expected growth rates generally
decrease with increasing $\Lambda_0$. In the limit of ideal MHD ($\Lambda_0>1$), 
there is negligible dependence on $Q$. In the resistive limit
($\Lambda_0<1$), however, growth rates decrease noticeably for $Q<1$. This shows
show that disk self-gravity affects the MRI through the background
equilibrium, even though density and potential perturbations are  
negligible (i.e. the linear response is
non-self-gravitating).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{MRI growth rates as a function of $Q$ for mid-planet
    Elsasser numbers $\Lambda_0=0.3$ (solid), $0.5$ (dotted), $1.0$
    (dashed) and $5.0$ (dash-dot).  
    \label{compare_growth_poly_uniresis}}
\end{figure}

\cite{sano99} found that for instability to occur the linear mode
wavelength $\lambda$ should fit inside the disk. That is,   
\begin{align}\label{sano_crit}
  \lambda \equiv
  \mathrm{max}\left(\lambda_\mathrm{ideal},\lambda_\mathrm{resis}\right)\lesssim
  2H, 
\end{align}
where the MRI wavelengths are given by 
\begin{align}\label{lambda_ideal}
  \frac{\lambda_\mathrm{ideal}}{2H} = \frac{4\pi}{\sqrt{15}} f \hat{v}_A =
  \frac{4\pi f}{\sqrt{15\beta\hat{\rho}}}
\end{align}
in the limit of ideal MHD, and 
\begin{align}\label{lambda_resis}
  \frac{\lambda_\mathrm{resis}}{2H} = \frac{2\pi}{\sqrt{3}}\frac{\hat{\eta}}{\hat{v}_A f} =
  \frac{2\pi f}{\Lambda_0}\sqrt{\frac{\hat{\rho}}{3\beta}} 
\end{align}
in the limit of high resistivity.

Since $\hat{\rho}$ is weakly dependent
on $Q$ (Fig. \ref{eqm_den}), self-gravity only affects the
MRI through the factor $f$, which increases with decreasing $Q$ (see
Fig. \ref{plot_fq} in Appendix \ref{appen1}). This suggests that
sufficiently strong self-gravity can stabilize the MRI by making
$ 2H<\lambda $.

In the ideal limit, we find $\lambda < 2H$ throughout most of the disk
for the values of $Q$ considered, so that self-gravity does not affect
growth rates significantly. However, the wavelength of the instability
increases. This is shown in Fig. \ref{compare_result_lambda10} which
plots the magnetic energies for $\Lambda_0=10$ and a range of Toomre $Q$
values. The number of vertical nodes decrease with $Q$, 
i.e. the disk accommodates fewer wavelengths because increasing
vertical self-gravity makes it thinner.

@@

@@

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not completely surpressed since
Eq. \ref{sano_crit} is only an approximation.) Although the function
$f$ does not change significantly for the range of $Q$ considered, the
dependence of $\lambda_\mathrm{resis}$ on $f$ (and therefore $Q$) is
amplified by the denominator $\Lambda_0<1$ in the resistive case. For
resistive cases we find no nodes in the magnetic energy $E_1$ except
at end points, i.e. only the longest wavelength survives.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/lambda_poly_uniresis}
  \caption{Approximate wavelengths of the most unstable resistive MRI modes as given by
    Eq. \ref{sano_crit}---\ref{lambda_resis}, normalized by the 
    disk thickness, as a function of height for a range of Toomre $Q$
    values.  
    \label{lambda_poly_resis}}
\end{figure}

@@

\subsubsection{Dependence on $k_x$}
The above experiments show that with increasing 
disk self-gravity, the MRI becomes more global in the vertical
direction. We find a similar result in the horizontal direction. 
In Fig. \ref{compare_growth_poly_kx} we plot MRI growth rates as a
function of $k_x$ for a range of $Q$ values. Increasing self-gravity
decreases the cut-off horizontal wavenumber for the MRI, implying that
small-scale MRI does not operate in massive disks.

@@

\subsection{Influence of self-gravity on the MRI through the linear
  response}  
Our goal here is to examine whether or not self-gravity can amplify
the density perturbations caused by the MRI. We choose a massive 
isothermal disk with $Q=0.2$ (corresponding to $Q_\mathrm{2D}=0.72$),
which is still expected to be marginally 
stable to gravitational instability \citep[][who find
a critical value of  $Q\simeq 0.2$]{mamat10}.   
The upper disk boundary is set to $Z_s=H$.

@@

\subsubsection{Ideal case}
We first consider the limit of ideal MHD by adopting a uniform
resistivity with $\Lambda_0=100$. %Non-zero density perturbations
%require $k_x\neq0$ so we consider linear modes as a function of $k_x$.   
Fig. \ref{gravity_energy} shows the mean gravitational potential
energy perturbation, as a function of $k_x$ for several values of
$\beta$. The potential perturbation is negligible for all cases when 
$k_xH\lesssim 0.5$, since the MRI becomes incompressible as $k_x\to
0$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_ideal}
  \caption{Average gravitational potential energy perturbation
    associated with the MRI, compared to the average magnetic energy
    perturbation, in a self-gravitating disk with $Q=0.2$ in 
    the limit of ideal MHD ($\Lambda_0=10^2$), for a range of field
    strengths $\beta$.  
    \label{gravity_energy}}
\end{figure}

For $\beta\gg 1$, i.e. a weak field, density perturbations are
negligible and the standard MRI operates. However, as $\beta$ is
lowered we find non-negligible potential perturbation for
$k_xH=O(1)$. This suggets that in a strongly magnetized disk that
still permits the MRI, the associated density perturbation can be
important when the disk is self-gravitating.

@@

\subsubsection{Resistive case}
We repeat the above calculation for resistive disks, but fix
$\beta=100$. Fig. \ref{gravity_energy_resis} shows the 
perturbed gravitational energy increases with   
resistivity. Interestingly, in the highly resistive case
the MRI can be energeically dominated by self-gravity, i.e. the case
with $\Lambda_0=0.1$ and $k_xH\simeq1.3$. We also calculated this mode
in the \emph{Cowling  
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally letting $Q\to\infty$ in Eq. \ref{lin_poisson}).
Fig. \ref{mri_massive_cowling} compares the magnetic energy in the
Cowling approximation to that of 
the full problem in . The growth rate increases when self-gravity is
included in the linear response, as expected, but only
slightly. $\gamma$ and $E_m$ are similar in both cases, indicating
that the instability in the self-gravitating calculation is 
still the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{Average gravitational potential energy perturbation
    associated with the MRI, compared to the average magnetic energy
    perturbation, in a self-gravitating disk with $Q=0.2$,
    for a range of midplane Elsasser numbers $\Lambda_0$ at fixed
    field strength $\beta=100$.    
    \label{gravity_energy_resis}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_cowling}
  \caption{Magnetic energy associated with the linear mode with
    largest gravitational-to-magnetic energy ratio in
    Fig. \ref{gravity_energy_resis} (solid) compared with that computed
    under the Cowling approximation (dotted). 
    \label{mri_massive_cowling}}
\end{figure}

An example is shown in
Fig. \ref{mri_massive_resis}, which  plots the energies assocaited with the 
linear MRI mode with $k_xH\sim1.3$ in the $\Lambda_0=0.1$  
disk. The gravitational potential energy exceeds the
magnetic energy near the midplane ($z\lesssim0.2H$). The growth rate $\gamma=0.25\Omega$
is not much smaller than that of the most unstable mode
($\gamma=0.36\Omega$) so significant density perturbations are
expected to grow on dynamical timescales.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_resis_sg}
  \caption{Example of a linear resistive MRI mode with significant
    gravitational potential energy perturbation.  The vertical line
    inidicates $\Lambda=1$.  
    \label{mri_massive_resis}}
\end{figure}

\subsubsection{Qualitative interpretation} 
In order to make sense of the above results, we return to ideal MHD. 
We consider regions close to the disk midplane ($z\sim 0$), where
self-gravity is expected to be most important. For this discussion we
will ignore stratification and set $d/dz\to \imgi k_z$ where $k_z$ is
a vertical wavenumber. The governing equations are then

\begin{align}
  &  0= v_A^2k^2\dvx + \imgi\sigma\left(\imgi\sigma \dvx - 2\Omega\dvy + \imgi k_x \w\right),\label{simp1}\\
  &  0= v_A^2k_z^2\left(\dvy + \frac{\imgi S
  }{\sigma}\dvx\right) + \imgi\sigma\left(\imgi\sigma\dvy +
  \frac{\kappa^2}{2\Omega}\dvx\right),\label{simp2}\\
  & 0 = -k_z^2\w + \frac{\sigma^2}{c_s^2}W + \sigma k_x \dvx, \label{simp3}\\
  & 0 = k^2\dphi + \frac{\Omega^2}{c_s^2Q}W,\label{simp4}
\end{align}
where $k^2 = k_z^2 + k_x^2$.

We imagine an iterative procedure to solve the above equations,
starting from the Cowling approximation. %defined to be the
%solution to Eq. \ref{simp1}---\ref{simp3} with potential perturbation
%set to zero. 
This is the standard MRI and we denote the solution as
$\dvx^{(0)}$, $\dvy^{(0)}$ and $W^{(0)}$.

We now include self-gravity. In general, the compressible MRI has
$W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{(0)} \to \dvx^{(0)} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
For the ideal MRI,  we assume $k_z\sim \Omega/v_A$. Then
$|\sigma^2/c_s^2k_z^2|\sim |\sigma^2/\Omega^2\beta|\ll1$ because
$|\sigma|\lesssim \Omega$ and we are considering $\beta\gtrsim
10$. Thus Eq. \ref{cond} is generally satisfied.

@@

@@

% not expect significant potential perturbations

%Conversely, for strong fields, only a single MRI wavelength occupies
%the fluid column.

The above arguments also apply to the resistive
limit. Eq. \ref{simp3}---\ref{simp4} are unchanged, while resistive
terms appearing in Eq. \ref{simp1}---\ref{simp2} only involve the
potential perturbation through $\w$. For the resistive MRI we assume
$k_z\sim v_A/\eta$ and $|\sigma|\sim v_A^2/\eta = \Lambda\Omega$
\citep{sano99}. Then $|\sigma^2/c_s^2k_z^2|\sim 1/\beta \ll 1$ so
Eq. \ref{cond} is satisfied. Noting that $k_z^2\sim
\Lambda^2\Omega^2\beta/c_s^2$, the feedback equation becomes
\begin{align}
  \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim
  \frac{1}{Q\left(f^2\hat{k}_x^2 + \beta\Lambda^2\right)},
\end{align}
so increasing the resistivity (decreasing $\Lambda$) should enhance
density perturbations.

>>>>

\section{MRI in self-gravitating disks}\label{result1}
In this section we focus on the MRI and use the vertical field setup
of case 1. We first consider MRI modes with negligble density/potential
perturbations to see the effect of self-gravity on the MRI through the
background stratification, then go on to examine MRI modes with
density/potential perturbations in massive disks.

\subsection{Influence of self-gravity on the MRI through the
  background equilibrium}
Here we use polytropic disks, so that the disk thickness is
well-defined. The upper disk boundary is set to 
$Z_s=0.99H$. We fix $\beta=100$ and $k_xH=0.1$ unless otherwise
stated.

@@

\subsubsection{Uniform
  resistivity}   
%We calculate the MRI as a function of $Q$ with uniform
%resistivity ($A=1$) for mid-plane Elsasser numbers $\Lambda_0\in[0.3,10]$.    
Fig. \ref{compare_growth_poly_uniresis} plots MRI growth rates as a
function of $Q$ and $\Lambda_0$. The resistivity is uniform ($A=1$). 
%As expected growth rates generally decrease with decreasing 
%$\Lambda_0$ \citep{jin96}. 
In the limit of ideal MHD ($\Lambda_0>1$), there is negligible
dependence on $Q$. In the resistive limit 
($\Lambda_0<1$), however, growth rates decrease noticeably for $Q<0.5$
($Q_\mathrm{2D}\lesssim 1.5$). This shows 
show that disk self-gravity affects the MRI through the background
equilibrium, even though density and potential perturbations are  
unimportant (i.e. the linear response is non-self-gravitating).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{MRI growth rates as a function of $Q$ for mid-planee 
    Elsasser numbers $\Lambda_0=0.3$ (solid), $0.5$ (dotted), $1.0$
    (dashed) and $5.0$ (dash-dot).  
    \label{compare_growth_poly_uniresis}}
\end{figure}

\cite{sano99} found that for MRI to operate, its 
wavelength $\lambda$ should fit inside the disk. That is,   
\begin{align}\label{sano_crit}
  \lambda \equiv
  \mathrm{max}\left(\lambda_\mathrm{ideal},\lambda_\mathrm{resis}\right)\lesssim
  2H, 
\end{align}
where the MRI wavelengths are given by 
\begin{align}\label{lambda_ideal}
  \frac{\lambda_\mathrm{ideal}}{2H} = \frac{4\pi}{\sqrt{15}} f \hat{v}_A =
  \frac{4\pi f}{\sqrt{15\beta\hat{\rho}}}
\end{align}
for ideal MHD, and 
\begin{align}\label{lambda_resis}
  \frac{\lambda_\mathrm{resis}}{2H} = \frac{2\pi}{\sqrt{3}}\frac{\hat{\eta}}{\hat{v}_A f} =
  \frac{2\pi f}{\Lambda_0}\sqrt{\frac{\hat{\rho}}{3\beta}} 
\end{align}
in the limit of high resistivity.

Since $\hat{\rho}$ is weakly dependent
on $Q$ (Fig. \ref{eqm_den}), self-gravity only affects the
MRI through the factor $f$, which increases with decreasing $Q$ (see
Fig. \ref{plot_fq} in Appendix \ref{appen1}).   
This implies that sufficiently strong self-gravity can stabilize the
MRI by making $ 2H<\lambda $.  %these perts are not gravitationally
                               %unstable

In the ideal limit, we find $\lambda < 2H$ throughout most of the disk
for the values of $Q$ considered, so self-gravity does not affect
growth rates significantly. However, the wavelength of the
instability, in units of $H$, increases. This is shown in
Fig. \ref{compare_result_lambda10} which plots the magnetic energies
for $\Lambda_0=10$ and a range of $Q$ values. The number of vertical
nodes decrease with $Q$, i.e. the disk accommodates fewer wavelengths
because increasing vertical self-gravity makes it thinner.

@@

%{\bf DISCUSS STRESS} Although for ideal MHD self-gravity makes no
%difference to the MRI growth rate, we find the associated stress does
%change with $Q$.

@@

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not completely surpressed since
Eq. \ref{sano_crit} is only an approximation.) Although the function
$f(Q)$ does not change significantly for the range of $Q$ considered, the
dependence of $\lambda_\mathrm{resis}$ on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. For resistive cases
we find no nodes in the magnetic energy $E_m$ except at end points,
i.e. only the longest wavelength survives.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/lambda_poly_uniresis}
  \caption{Approximate wavelengths of the most unstable MRI modes as given by
    Eq. \ref{sano_crit}---\ref{lambda_resis}, normalized by the 
    disk thickness, as a function of height. MRI is expected to
    operate if $\lambda/2H\lesssim 1$. 
    \label{lambda_poly_resis}}
\end{figure}

@@

\subsubsection{Dependence on $k_x$}
The above experiments show that with increasing 
disk self-gravity, the MRI becomes more global in the vertical
direction. We find a similar result in the horizontal direction. 
In Fig. \ref{compare_growth_poly_kx} we plot MRI growth rates as a
function of $k_x$ for a range of $Q$ values. Increasing self-gravity
decreases the cut-off horizontal wavenumber for the MRI. We checked
that these modes have negligible density perturbations. Then we can 
understand this result by invoking the instability criteria for  
incompressible MRI in an unstratified Keplerian disk,
\begin{align}
  v_A^2(k_z^2 + k_x^2) < 3\Omega^2,
\end{align}
where $k_z$ is a vertical wavenumber \citep{kim00}. Setting $k_z^2\sim
\Omega^2/v_A^2$ and non-dimensionalizing, we find
\begin{align} 
  k_xH \lesssim \frac{\sqrt{\beta}}{f},
\end{align}
where order-unity factors have been dropped. Despite a simplistic
approach, this demonstrates that with increasing self-gravity
(increasing $f$), we expect that MRI modes with small radial
lengthscales to be suppressed.

%, implying that
%small-scale MRI does not operate in massive disks.

@@

\subsection{Influence of self-gravity on the MRI through the linear
  response}  
Our goal here is to examine whether or not self-gravity can amplify
the density perturbations caused by the MRI. We compute unstable modes
in a massive isothermal disk with $Q=0.2$ (corresponding to
$Q_\mathrm{2D}=0.72$), which is still expected to be marginally  
stable to gravitational instability \citep[][who find
a critical value of  $Q\simeq 0.2$]{mamat10}.   
The upper disk boundary is set to $Z_s=H$.

@@

\subsubsection{Ideal disks}
We first consider ideal MHD by adopting a uniform 
resistivity with $\Lambda_0=100$. %Non-zero density perturbations
%require $k_x\neq0$ so we consider linear modes as a function of $k_x$.   
Fig. \ref{gravity_energy} plots MRI growth rates as a function of
$k_x$ for several values of $\beta$. The curves are color-coded
according to $\tau$. The potential perturbation is
negligible for all cases when $k_xH\lesssim 0.5$, since the MRI
becomes incompressible as $k_x\to 0$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_ideal}
  \caption{Growth rates of MRI modes in an isothermal self-gravitating
    disk with $Q=0.2$ ($Q_\mathrm{2D}=0.72$) in the limit of ideal MHD
    ($\Lambda_0=10^2$), for a range of field strengths $\beta$. The
    colorbar measures the importance of self-gravity by $\tau$.  
    %{\bf ADD BETA=8}
    \label{gravity_energy}}
\end{figure}

For $\beta\gg 1$, i.e. a weak field, density perturbations are
negligible and the standard MRI operates. However, as $\beta$ is
lowered and the MRI growth rate reduced, we find non-negligible
potential perturbation for $k_xH=O(1)$. This suggets that in a
strongly magnetized disk that still permits the MRI, the associated
density perturbation can be important when the disk is
self-gravitating. %but formall stable

%critical beta, below which MRI is surpressed, increases with the
%strength of self-gravity.

@@

\subsubsection{Resistive disks}
We repeat the above calculation for resistive disks, but fix
$\beta=100$ and vary the mid-plane Elasser number $\Lambda_0$. Growth
rates are shown in Fig. \ref{gravity_energy_resis}. 
Interestingly, in the highly resistive case
the magnetic and gravitational energies can be comparable. For 
$\Lambda_0=0.1$ and $k_xH\simeq1.3$, $\tau\sim 0.3$ which corresponds
to $\avg{E_g}\sim 0.5\avg{E_m}$.  
Fig. \ref{mri_massive_cowling}, we compare the magnetic
energy of this mode to that computed in in the \emph{Cowling  
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally letting $Q\to\infty$ in Eq. \ref{lin_poisson}).  
The growth rate increases slightly when self-gravity is included in the linear
response, since self-gravity is usually destabilizing. However, the
similarity between $\gamma$ and $E_m$ indicates that the instability
in the self-gravitating calculation is fundamentally the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{
    Growth rates of MRI modes in an isothermal self-gravitating
    disk with $Q=0.2$ ($Q_\mathrm{2D}=0.72$) at fixed $\beta=100$, 
    for a range of mid-plane Elasser numbers. The resistivity is
    uniform. The colorbar measures the importance of self-gravity by $\tau$. 
    \label{gravity_energy_resis}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_cowling}
  \caption{Magnetic energy associated with the linear mode with
    largest gravitational-to-magnetic energy ratio in
    Fig. \ref{gravity_energy_resis} (solid) compared with that computed
    under the Cowling approximation (dotted). %{\bf REDO FIGURE}
    \label{mri_massive_cowling}}
\end{figure}

Fig. \ref{mri_massive_resis} plots the energies assocaited with the 
MRI mode discussed above. 
%with $k_xH\sim1.3$ in the $\Lambda_0=0.1$  
%disk. 
The gravitational energy exceeds the 
magnetic energy near the midplane ($z\lesssim0.2H$). The growth rate $\gamma=0.25\Omega$
is not much smaller than that of the most unstable mode
($\gamma=0.36\Omega$ for $k_xH=0.1$), so significant density
perturbations are expected to grow on dynamical timescales for this
system, even though GI should not develop.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_resis_sg}
  \caption{Example of a resistive MRI mode with significant
    gravitational potential perturbation.  The vertical line
    inidicates $\Lambda=1$.  
    \label{mri_massive_resis}}
\end{figure}

\subsubsection{Qualitative interpretation} 
To make sense of the above results, we first return to ideal MHD and 
consider regions close to the disk midplane ($z\sim 0$), where
self-gravity is expected to be most important. For this discussion we
will ignore stratification and set $d^2/dz^2\to -k_z^2$. The governing
equations are then

\begin{align}
  &  0= v_A^2k^2\dvx + \imgi\sigma\left(\imgi\sigma \dvx - 2\Omega\dvy + \imgi k_x \w\right),\label{simp1}\\
  &  0= v_A^2k_z^2\left(\dvy + \frac{\imgi S
  }{\sigma}\dvx\right) + \imgi\sigma\left(\imgi\sigma\dvy +
  \frac{\kappa^2}{2\Omega}\dvx\right),\label{simp2}\\
  & 0 = -k_z^2\w + \frac{\sigma^2}{c_s^2}W + \sigma k_x \dvx, \label{simp3}\\
  & 0 = k^2\dphi + \frac{\Omega^2}{c_s^2Q}W,\label{simp4}
\end{align}
where $k^2 = k_z^2 + k_x^2$. We imagine an iterative procedure to
solve the above equations, starting from the Cowling approximation. 
%defined to be the
%solution to Eq. \ref{simp1}---\ref{simp3} with potential perturbation
%set to zero. 
This is the standard MRI and we denote the solution as
$\dvx^{(0)}$, $\dvy^{(0)}$ and $W^{(0)}$.

We now include self-gravity. In general, the compressible MRI has
$W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a local positive (negative) density
perturbation causes a negative (positive) local potential
perturbation. We then insert $\dphi$ back into the momentum and
continuity equations, and ask how does this potential perturbation
modify the Cowling solution? Writing $\dvx^{(0)} \to \dvx^{(0)} +
\dvx^{(1)}$ and similarly for $\dvy $ and $ W$, we find 
%regard $\dphi$ to alter the Cowling solution, so
%that 
%Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
For the ideal MRI,  we assume $k_z\sim \Omega/v_A$. Then
$|\sigma^2/c_s^2k_z^2|\sim |\sigma^2/\Omega^2\beta|\ll1$ because
$|\sigma|\lesssim \Omega$ and we are considering $\beta\gtrsim 10$. 
Thus Eq. \ref{cond} is generally satisfied.

@@

% not expect significant potential perturbations 
%Conversely, for strong fields, only a single MRI wavelength occupies
%the fluid column.

 The above arguments can be adapted to the resistive
 limit. Eq. \ref{simp3}---\ref{simp4} are unchanged, while resistive
 terms appearing in Eq. \ref{simp1}---\ref{simp2} only involve the
 potential perturbation through $\w$. For the resistive MRI we assume
 $k_z\sim v_A/\eta$ and $|\sigma|\sim v_A^2/\eta = \Lambda\Omega$
 \citep{sano99}. Then $|\sigma^2/c_s^2k_z^2|\sim 1/\beta \ll 1$ so
 Eq. \ref{cond} is satisfied. Noting that $k_z^2\sim
 \Lambda^2\Omega^2\beta/c_s^2$, the feedback equation becomes
 \begin{align}
   \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim
   \frac{1}{Q\left(f^2\hat{k}_x^2 + \beta\Lambda^2\right)},
 \end{align}
 so increasing the resistivity (decreasing $\Lambda$) should enhance
 density perturbations.

@@
